Comparing Early Evolution of Flames in X-ray Bursts in Two and Three Dimensions

### 2023-03-24 xrb3/paper.tex
spelling and grammar

Moving to 3D is very expensive, so to further reduce the computational
expense, we use an aniostropic resolution, with 32 cm resolution
laterally and 16 cm resolution vertically.  This is finer vertically
than done in \citet{harpole:2021}.  Because the hotspot is placed in
the center of the domain, the size of the domain also becomes a
constraint, and as a result, the evolution time we can reach will
become limited by the time it takes the burning to approach the edge
of the domain.  Consequently, we focus here on the early evolution.

@@

The simulations were run on the OLCF Summit supercomputer, on 342 to
1366 nodes, with 6 NVIDIA V100 GPUs per node.  The entire computation
was offloaded to GPUs using the \amrex\ C++ lambda-capturing mechanism, as described in
\citet{castro_gpu}.  GPU-offloading was critical to being able to
perform these simulations, since we run more than an order of
magnitude faster using all the GPUs on a node as compared to all of
the CPUs on the node.  Overall, about 250,000 node-hours were used for
the calculation.  The full state of the calculation was output only
every 0.005 s, as each output file is 2.4 TB in size (for single
precision data).  We output a few fields more frequently for
visualization, and global diagnostics were output every timestep.  We
ran a 2D simulation with the same resolution and refinement strategy
to compare with here.

@@

Because the 3D flame does not stay perfectly circular, it is difficult
to define the flame speed using the same method employed in
\citet{eiden:2020}.  Instead, we will look at the mass of the ash
material to assess how quickly the burning is taking place.
Figure~\ref{fig:mass_plot} shows the mass of the different species as
a function of time, for the 2D and 3D calculation.  The axisymmetric
2D calculation has slightly less volume than the 3D domain, since
rotating about the symmetry axis produces a cylindrical domain
inscribed in the 3D domain.  To compensate for this, we scale the
masses by the total mass on the grid.  This plot shows that the
masses of the heavy species, especially \isot{C}{12}, grow quickly
with time.  The differences between the 2D and 3D simulations appear
small---the burning is slightly faster in 2D, but the trends track
very well.  This slight difference is likely because the assumption of
axisymmetry there does not allow for the complex structure we see in
the evolution of composition in the 3D flame.  This strong agreement
suggests that using 2D axisymmetric calculations is a good model for
the early flame evolution.  And since they are so much less
computationally expensive, we this will allow us to explore much more
complex reaction networks and understand the nucleosynthesis in more
detail.

@@

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernovae, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\citep{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accurate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in \castro~\citep{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.

>>>>

Moving to 3D is very expensive, so to further reduce the computational
expense, we use an anisotropic resolution, with 32 cm resolution
laterally and 16 cm resolution vertically.  This is finer vertically
than done in \citet{harpole:2021}.  Because the hotspot is placed in
the center of the domain, the size of the domain also becomes a
constraint, and as a result, the evolution time we can reach will
become limited by the time it takes the burning to approach the edge
of the domain.  Consequently, we focus here on the early evolution.

@@

The simulations were run on the OLCF Summit supercomputer, on 342 to
1366 nodes, with 6 NVIDIA V100 GPUs per node.  The entire computation
was offloaded to GPUs using the \amrex\ C++ abstraction layer for parallel \texttt{for} loops, as described in
\citet{castro_gpu}.  GPU offloading was critical to being able to
perform these simulations, since we run more than an order of
magnitude faster using all the GPUs on a node as compared to all of
the CPU cores on the node.  Overall, about 250,000 node-hours were used for
the calculation.  The full state of the calculation was output only
every 0.005 s, as each output file is 2.4 TB in size (for single
precision data).  We output a few fields more frequently for
visualization, and global diagnostics were output every timestep.  We
ran a 2D simulation with the same resolution and refinement strategy
to compare with here.

@@

Because the 3D flame does not stay perfectly circular, it is difficult
to define the flame speed using the same method employed in
\citet{eiden:2020}.  Instead, we will look at the mass of the ash
material to assess how quickly the burning is taking place.
Figure~\ref{fig:mass_plot} shows the mass of the different species as
a function of time, for the 2D and 3D calculation.  The axisymmetric
2D calculation has slightly less volume than the 3D domain, since
rotating about the symmetry axis produces a cylindrical domain
inscribed in the 3D domain.  To compensate for this, we scale the
masses by the total mass on the grid.  This plot shows that the
masses of the heavy species, especially \isot{C}{12}, grow quickly
with time.  The differences between the 2D and 3D simulations appear
small---the burning is slightly faster in 2D, but the trends track
very well.  This slight difference is likely because the assumption of
axisymmetry there does not allow for the complex structure we see in
the evolution of composition in the 3D flame.  This strong agreement
suggests that using 2D axisymmetric calculations is a good model for
the early flame evolution.  And since they are so much less
computationally expensive, this will allow us to explore much more
complex reaction networks and understand the nucleosynthesis in more
detail.

@@

There is still a role for three-dimensional simulations---the
convective burning in the accreted layer should create turbulence that
the flame will encounter as it propagates.  Although we have shown
that for the flame in a Type Ia supernova, the turbulence from the
era of convection is not strong enough to disrupt the
flame~\citep{wdturb}, the flame in an XRB is considerably slower and
thicker, so it remains to be seen what effect it might have.  This can
be assessed in a different geometry, like a long channel, which would
allow us to trade domain size for resolution to increase the numerical
Reynolds number.  We also will consider higher-order simulation
methodologies which could allow us to capture these effects at lower
resolution.  We've already demonstrated a fully fourth-order accurate
(in space and time) algorithm for coupling hydro, diffusion, and
reactions in \castro~\citep{castro-sdc} that could be used here.  The
main outstanding issue with applying that work to the present problem
is the multilevel time integration.
